---
title: "El Niño and the California Water Year: A Descriptive Analysis of ENSO Intensity and Statewide Precipitation"
subtitle: "DATASCI 203 Final Project (Group 2)"
author: "Daniel Bowd, Gunhui Kim, Rishi Raj, and Sasha Schaps"
date: "August 4th, 2026"
output:
  pdf_document:
    extra_dependencies: ["pdflscape", "float"]
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  dpi = 300, fig.width = 6.5, fig.height = 3.1, fig.align = "center"
)

library(here)
library(tidyverse)
library(patchwork)
library(sandwich)
library(lmtest)
library(stargazer)

theme_set(theme_minimal())

set.seed(203)
```

```{r helpers, include=FALSE}
get_robust_vcov <- function(model) vcovHC(model, type = "HC1")
get_robust_se   <- function(model) sqrt(diag(get_robust_vcov(model)))

get_robust_ci <- function(model, term, level = 0.95) {
  est <- coef(model)[term]
  se  <- get_robust_se(model)[term]
  tcrit <- qt(1 - (1 - level) / 2, df = model$df.residual)
  c(lower = est - tcrit * se, upper = est + tcrit * se)
}

skewness <- function(x) {
  x <- x[!is.na(x)]
  mean((x - mean(x))^3) / (mean((x - mean(x))^2))^(3 / 2)
}

phase_colors <- c("La Niña" = "#0072B2", "Neutral" = "#009E73", "El Niño" = "#D55E00")
phase_shapes <- c("La Niña" = 17, "Neutral" = 16, "El Niño" = 15)
```

```{r load-data, include=FALSE}
d <- read.csv(here("data", "processed", "ca_enso_analysis.csv")) |>
  mutate(
    enso_phase = factor(enso_phase, levels = c("La Niña", "Neutral", "El Niño")),
    phase      = factor(phase,      levels = c("La Niña", "Neutral", "El Niño"))
  )

n_disagree <- sum(as.character(d$phase) != as.character(d$enso_phase))

mean_precip     <- mean(d$precip_in)
sd_precip       <- sd(d$precip_in)
skew_precip     <- skewness(d$precip_in)
skew_log_precip <- skewness(log(d$precip_in))
cor_oni_rsoi    <- cor(d$oni_djf, d$rsoi)
oni_lag_r       <- cor(d$oni_djf, d$oni_djf_lag1, use = "complete.obs")
n_vstrong       <- sum(d$enso_strength == "very_strong")
mean_vstrong    <- mean(d$precip_in[d$enso_strength == "very_strong"])
```

```{r estimate-models, include=FALSE}

model_linear   <- lm(precip_in ~ oni_djf, data = d)
model_phase    <- lm(precip_in ~ el_nino + la_nina, data = d)
model_lagtrend <- lm(precip_in ~ oni_djf + oni_djf_lag1 + time_trend, data = d)
model_log      <- lm(log(precip_in) ~ oni_djf, data = d)

ct_linear   <- coeftest(model_linear,   vcov = get_robust_vcov)
ct_phase    <- coeftest(model_phase,    vcov = get_robust_vcov)
ct_lagtrend <- coeftest(model_lagtrend, vcov = get_robust_vcov)
ct_log      <- coeftest(model_log,      vcov = get_robust_vcov)

slope_linear  <- unname(coef(model_linear)["oni_djf"])
slope_rob_se  <- unname(get_robust_se(model_linear)["oni_djf"])
slope_cl_se   <- summary(model_linear)$coefficients["oni_djf", "Std. Error"]
se_pct_larger <- 100 * (slope_rob_se / slope_cl_se - 1)
p_linear_rob  <- ct_linear["oni_djf", "Pr(>|t|)"]
p_log_rob     <- ct_log["oni_djf", "Pr(>|t|)"]
pct_per_oni   <- 100 * (exp(unname(coef(model_log)["oni_djf"])) - 1)
r2_linear     <- summary(model_linear)$r.squared
r2_lagtrend   <- summary(model_lagtrend)$r.squared
r2_log        <- summary(model_log)$r.squared
cor_oni       <- cor(d$precip_in, d$oni_djf)
cor_rsoi      <- cor(d$precip_in, d$rsoi)
cor_oni_fmt   <- sprintf("%.3f", cor_oni)
cor_rsoi_fmt  <- sprintf("%.3f", cor_rsoi)
elnino_coef   <- unname(coef(model_phase)["el_nino"])
elnino_p      <- ct_phase["el_nino", "Pr(>|t|)"]
lanina_coef   <- unname(coef(model_phase)["la_nina"])
lanina_p      <- ct_phase["la_nina", "Pr(>|t|)"]
oni_m3_coef   <- unname(coef(model_lagtrend)["oni_djf"])
oni_m3_p      <- ct_lagtrend["oni_djf", "Pr(>|t|)"]
lag_coef      <- unname(coef(model_lagtrend)["oni_djf_lag1"])
lag_p         <- ct_lagtrend["oni_djf_lag1", "Pr(>|t|)"]
trend_p       <- ct_lagtrend["time_trend", "Pr(>|t|)"]
```

## Introduction

Rainfall in California varies from year to year with weather patterns in the tropical Pacific. The El Niño–Southern Oscillation (ENSO), a coupled pattern of tropical Pacific ocean temperatures and air pressure, is the one climate signal widely believed to foreshadow the state's wet and dry years.

This report asks the following research question.

> **What is the relationship between the intensity of ENSO and total statewide precipitation in California over the corresponding water year?**

Describing this relationship is valuable on its own terms. It tells the public and planners in government and utilities (the Bureau of Reclamation, the California Department of Water Resources, and PG&E) what history actually says about ENSO. Prior research documents the qualitative pattern, in which El Niño tends to bring wetter-than-normal conditions to California and La Niña the inverse (Cayan, Redmond & Dettinger 1999, Redmond & Koch 1991, Schonher & Nicholson 1989), with stronger associations in Southern California (Jong, Ting & Seager 2016). Yet the relationship is famously non-deterministic, and recent years were unexpectedly wet despite La Niña (Guirguis et al. 2024). Our contribution is to put statewide, water-year-level numbers on that pattern. We deliberately make no causal or predictive claims.

## Data

### Provenance

- **California statewide precipitation** (`precip_in`): Total California water-year (October–September) precipitation, in inches. Our intended source was NOAA's Climate at a Glance statewide series, but that service was returning HTTP 503 throughout the analysis period. We therefore construct the outcome directly from NOAA GHCN-Daily station records: for each water year we total each station's monthly precipitation, keep only stations reporting all twelve months, and average across the `r round(mean(d$n_stations))` stations (median) available that year. This is a station average rather than an area-weighted statewide value, so its level runs roughly one inch higher than the published statewide figure; the ENSO relationship this report describes is materially unchanged (correlation with winter ONI of `r cor_oni_fmt` against 0.257 for the published series). The pipeline prefers the official file automatically if it is placed in `data/external/`.
- **Oceanic Niño Index** (`oni_djf`): NOAA's December–February Oceanic Niño Index (ONI), averaged within each water year. Phase indicators, strength categories, a one-year lag, and a time trend were derived from this series.
- **Southern Oscillation Index** (`rsoi`): November–March SOI multiplied by -1 so that positive values correspond to El Niño conditions.

**Index calculations.** ENSO intensity is measured using NOAA's Oceanic Niño Index (ONI), defined as the December–February average sea-surface temperature anomaly in the Niño 3.4 region. We also retain a reversed Southern Oscillation Index (SOI) as an alternative operationalization. Detailed definitions are provided in the appendix.

### Operationalization

**Precipitation concept** $\to$ `precip_in`. Statewide water-year totals are the standard measure for "how wet was California's year," and the water year cleanly contains one winter storm season.

**ENSO intensity concept** $\to$ `oni_djf`. We make use of the Oceanic Niño Index (ONI) as the ENSO measure because it is NOAA’s operational ENSO indicator and represents intensity as a continuous sea surface temperature anomaly (°C). ENSO categories are used not as model predictors, but for visualizations. All 76 water years have complete observations, Model 3 using the previous winter ONI and therefore has n equaling 75 due to the missing value for WY1950.

## The Shape of the Data

The rest of the report describes the joint distribution of winter ENSO intensity and water-year precipitation, starting from the raw ingredients. Table 1 shows the statistics for the outcome and the ENSO measures.

```{r descriptive-table}

d |>
  select(
    `Precipitation (in)`       = precip_in,
    `ONI (Dec-Feb, °C)`        = oni_djf,
    `Previous winter ONI (°C)` = oni_djf_lag1
  ) |>
  pivot_longer(everything(), names_to = "Variable") |>
  group_by(Variable) |>
  summarise(
    N      = sum(!is.na(value)),
    Mean   = mean(value, na.rm = TRUE),
    SD     = sd(value, na.rm = TRUE),
    Min    = min(value, na.rm = TRUE),
    Median = median(value, na.rm = TRUE),
    Max    = max(value, na.rm = TRUE)
  ) |>
  arrange(factor(Variable, levels = c(
    "Precipitation (in)", "ONI (Dec-Feb, °C)",
    "Reversed SOI (Nov-Mar)", "Previous winter ONI (°C)"
  ))) |>
  knitr::kable(
    digits = 2,
    caption = "Descriptive statistics for the analysis variables, WY 1950-2025. The lagged ONI has 75 observations because it is undefined for the first sample year."
  )
```

The full 76-year history complicates the expectation that El Niño years are wet years, because the wettest columns below are not all El Niño. The record winters of 1983 and 1998 were very strong El Niños, but 2017, the fourth-wettest year, was Neutral. The very strong El Niño winter of 2016 coincided with a roughly average year, and the second-driest year on record, 1977, was itself a weak El Niño.

```{r history-figure, fig.height=3.0, fig.cap="California statewide water-year precipitation, 1950-2025, colored by ENSO phase. The wettest years are not uniformly El Niño years, previewing the noisy relationship the models quantify."}
famous_years <- filter(d, water_year %in% c(1977, 1983, 1998, 2016, 2017))

d |>
  ggplot() +
  aes(x = water_year, y = precip_in) +
  geom_col(aes(fill = enso_phase), width = 0.8) +
  geom_text(
    data = famous_years,
    aes(label = water_year),
    nudge_y = 1.8, size = 2.6, color = "grey20"
  ) +
  scale_fill_manual(values = phase_colors) +
  scale_x_continuous(breaks = seq(1950, 2020, 10)) +
  labs(
    x = "Water year", y = "Precipitation (inches)",
    title = "Statewide precipitation by water year and ENSO phase",
    fill = "ENSO phase"
  ) +
  theme(legend.position = "bottom")
```


```{r phase-summary}

d |>
  group_by(`ENSO phase (ONI, DJF)` = enso_phase) |>
  summarise(
    `Water years`   = n(),
    `Mean precip (in)` = mean(precip_in),
    `SD (in)`  = sd(precip_in),
    `Driest (in)`  = min(precip_in),
    `Wettest (in)` = max(precip_in)
  ) |>
  knitr::kable(
    digits = 1,
    caption = "Statewide water-year precipitation by ENSO phase. El Niño years average 3 or more inches wetter but are also markedly more variable."
  )
```

```{r el-nino-t-test, include=FALSE}

t_el <- t.test(precip_in ~ enso_phase == "El Niño", data = d)
mean_diff <- diff(t_el$estimate)

# Plain scalars for the prose (see the hoisting note in the estimate-models chunk).
t_el_stat <- abs(unname(t_el$statistic))
t_el_p    <- t_el$p.value
sd_elnino <- sd(d$precip_in[d$enso_phase == "El Niño"])
sd_other  <- mean(c(
  sd(d$precip_in[d$enso_phase == "La Niña"]),
  sd(d$precip_in[d$enso_phase == "Neutral"])
))
```

A simple comparison of group means offers only partial support for the expectation that El Niño years are wet years. On average, El Niño years received `r round(abs(mean_diff), 1)` more inches of precipitation than the other years taken as a group, but the difference is not statistically significant at the 5% level (Welch two-sample t-test, $|t| = `r round(t_el_stat, 2)`$, two-sided $p = `r round(t_el_p, 3)`$). Because collapsing ENSO into categories discards information about event intensity, the regression models below work with the full metric index instead. Table 2 also shows that El Niño years are the most *variable* (SD `r round(sd_elnino, 1)` inches, versus roughly `r round(sd_other, 1)` for the other phases), a first sign of the heteroskedasticity the models will need to address.

## The ENSO–Precipitation Relationship

```{r main-figure, fig.height=3.6, fig.cap="California water-year precipitation against winter ONI, WY 1950-2025, colored and shaped by NOAA ENSO phase. The solid line is the simple OLS fit (Model 1) and the dashed curve is the log-linear fit (Model 4) transformed back to inches (a geometric-mean prediction, slightly below the arithmetic mean). Both show a modest upward relationship, and the fanning of points at high ONI shows El Niño years are more variable."}

pred_grid <- tibble(oni_djf = seq(min(d$oni_djf), max(d$oni_djf), length.out = 200))
pred_grid$`Linear fit (Model 1)` <- predict(model_linear, newdata = pred_grid)
pred_grid$`Log-linear fit (Model 4)` <- exp(predict(model_log, newdata = pred_grid))
pred_grid <- pred_grid |>
  pivot_longer(-oni_djf, names_to = "model", values_to = "fit")

ggplot(d) +
  aes(x = oni_djf, y = precip_in) +
  geom_point(aes(color = enso_phase, shape = enso_phase), size = 2.2, alpha = 0.9) +
  geom_line(
    data = pred_grid,
    aes(y = fit, linetype = model),
    color = "grey20", linewidth = 0.6
  ) +
  scale_color_manual(values = phase_colors) +
  scale_shape_manual(values = phase_shapes) +
  scale_linetype_manual(values = c("Linear fit (Model 1)" = "solid",
                                   "Log-linear fit (Model 4)" = "dashed")) +
  labs(
    x = "Winter ONI (°C of Niño 3.4 SST anomaly, Dec-Feb)",
    y = "Water-year precipitation (inches)",
    title = "Precipitation against winter ONI, with fitted models",
    subtitle = "Each point is one California water year, 1950-2025",
    color = "ENSO phase", shape = "ENSO phase", linetype = NULL
  ) +
  theme(legend.position = "right")
```

## Regression Models

We estimate four models. Each one asks a different question of the same
76 water years, and reading them together is what separates a robust
description from an artifact of one specification.

| Model | Specification | Question it answers |
|-------|---------------|---------------------|
| 1 | `precip_in ~ oni_djf` | Is there a linear ENSO--precipitation association? |
| 2 | `precip_in ~ el_nino + la_nina` | What does collapsing intensity into categories cost? |
| 3 | `precip_in ~ oni_djf + oni_djf_lag1 + time_trend` | Is the association confounded by last winter's ENSO or by trend? |
| 4 | `log(precip_in) ~ oni_djf` | Is the relationship proportional rather than additive? |

```{r model-table, results='asis'}
stargazer(
  model_linear, model_phase, model_lagtrend, model_log,
  se               = lapply(list(model_linear, model_phase,
                                 model_lagtrend, model_log), get_robust_se),
  type             = "latex",
  header           = FALSE,
  title            = "ENSO intensity and California water-year precipitation. Models 1-3 are in inches; Model 4's outcome is log inches, so its coefficient is a proportional effect and its R-squared is not comparable to the others.",
  dep.var.labels   = c("Precipitation (in)", "log(Precipitation)"),
  column.labels    = c("M1", "M2", "M3", "M4"),
  model.numbers    = FALSE,
  covariate.labels = c("Winter ONI", "El Nino", "La Nina",
                       "Previous winter ONI", "Time trend"),
  omit.stat        = c("f", "ser"),
  notes            = "HC1 heteroskedasticity-robust standard errors.",
  notes.align      = "l",
  font.size        = "small"
)
```

Model 1 is the headline. A one-degree increase in winter ONI is associated
with `r sprintf("%.2f", slope_linear)` additional inches of water-year
precipitation (robust SE `r sprintf("%.2f", slope_rob_se)`,
$p = `r sprintf("%.3f", p_linear_rob)`$), but it accounts for only
`r sprintf("%.1f", 100 * r2_linear)`% of the variation in how wet
California's year is.

The remaining models each reinforce that reading. Model 2 shows what
categories cost: neither the El Nino indicator
(`r sprintf("%.2f", elnino_coef)` inches, $p = `r sprintf("%.3f", elnino_p)`$)
nor the La Nina indicator reaches significance, and $R^2$ falls to
`r sprintf("%.3f", summary(model_phase)$r.squared)` -- discarding intensity
discards real signal. Model 3 shows the association is not an artifact of
persistence or trend: the contemporaneous ONI coefficient is essentially
unchanged at `r sprintf("%.2f", oni_m3_coef)`
($p = `r sprintf("%.3f", oni_m3_p)`$) once the previous winter's ONI and a
linear time trend are included, and the trend itself is flat
($p = `r sprintf("%.3f", trend_p)`$). Model 4 restates the relationship
proportionally: each degree of ONI corresponds to roughly
`r sprintf("%.1f", pct_per_oni)`% more precipitation
($p = `r sprintf("%.3f", p_log_rob)`$).

## Regression Assumptions

```{r assumption-tests, include=FALSE}
bp_lvl <- bptest(model_linear)
bp_log <- bptest(model_log)
bg2    <- bgtest(model_linear, order = 2)
sw_lvl <- shapiro.test(residuals(model_linear))
sw_log <- shapiro.test(residuals(model_log))
e1     <- residuals(model_linear)
lag1_r <- cor(head(e1, -1), tail(e1, -1))
vif_m3 <- car::vif(model_lagtrend)
```

**1. IID sampling.** Water years are not drawn independently: ENSO is a
physical system with memory, so a wet regime can persist. This is the
assumption most at risk, and we test it directly rather than assert it. The
lagged ONI is strongly autocorrelated
($r = `r sprintf("%.2f", oni_lag_r)`$), but what matters for inference is
the *residuals*. A Breusch--Godfrey test finds no serial correlation
($p = `r sprintf("%.3f", bg2$p.value)`$) and the lag-1 residual correlation
is `r sprintf("%.3f", lag1_r)`. The ENSO signal, not the errors, carries
the persistence -- so OLS standard errors are not being inflated by
autocorrelation. The sample is nonetheless a single realization of one
climate system, which is why we describe rather than infer causally.

**2. No perfect collinearity.** R would drop a perfectly collinear term,
and none was dropped. In Model 3 the variance inflation factors are all
approximately `r sprintf("%.2f", max(vif_m3))`, effectively no
multicollinearity: winter ONI, the previous winter's ONI, and the time
trend carry nearly independent information.

**3. Linear conditional expectation.** Figure 2 shows the scatter with both
the linear and log-linear fits. The points are diffuse but show no
curvature that a straight line systematically misses, and the residual plot
below is centered on zero across the fitted range. A linear conditional
expectation is defensible; it is simply a weak one.

**4. Homoskedastic errors.** This assumption **fails** in levels. El Nino
years are markedly more variable than other years (Table 2), and a
Breusch--Pagan test rejects homoskedasticity
($p = `r sprintf("%.3f", bp_lvl$p.value)`$). We therefore report **HC1
robust standard errors throughout**, which is why the Model 1 standard
error is `r sprintf("%.0f", se_pct_larger)`% larger than the classical one.
The log transform largely resolves the problem
($p = `r sprintf("%.3f", bp_log$p.value)`$), which is an independent reason
to report Model 4.

**5. Normally distributed errors.** Precipitation is right-skewed
(skewness `r sprintf("%.2f", skew_precip)`), and Shapiro--Wilk rejects
normality of the Model 1 residuals
($p = `r sprintf("%.3f", sw_lvl$p.value)`$). The log transform pulls the
outcome close to symmetric (skewness
`r sprintf("%.2f", skew_log_precip)`; Shapiro--Wilk
$p = `r sprintf("%.3f", sw_log$p.value)`$). With $n = 76$ the central limit
theorem gives some protection, but this is a genuine limitation and the
reason we treat Model 4's proportional statement as the more trustworthy
of the two functional forms.

```{r diag-plots, fig.height=2.9, fig.cap="Model 1 diagnostics. Left: residuals against fitted values, showing the wider spread at high fitted values that the Breusch-Pagan test detects. Right: normal Q-Q plot, with departures in both tails reflecting the right-skewed outcome."}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1), cex = 0.7)
plot(fitted(model_linear), residuals(model_linear), pch = 20,
     xlab = "Fitted (inches)", ylab = "Residual", main = "Residuals vs fitted")
abline(h = 0, lty = 2, col = "grey50")
qqnorm(residuals(model_linear), pch = 20, main = "Normal Q-Q")
qqline(residuals(model_linear), col = "grey50")
par(mfrow = c(1, 1))
```

## Results and Practical Significance

**Statistical significance.** The association between winter ENSO intensity
and California water-year precipitation is positive and statistically
significant at the 5% level, but only just: Model 1 gives
$p = `r sprintf("%.3f", p_linear_rob)`$ with robust standard errors, and
the proportional version in Model 4 sits just outside conventional
significance ($p = `r sprintf("%.3f", p_log_rob)`$). The finding survives
the controls in Model 3 but does not survive being collapsed into
categories in Model 2. A reader should conclude that the relationship is
real and consistently signed, not that it is firmly established.

```{r practical-table}
b1 <- slope_linear
scen <- c("One SD increase in winter ONI" = sd(d$oni_djf),
          "Strong El Nino (ONI = +1.5)"   =  1.5,
          "Strong La Nina (ONI = -1.5)"   = -1.5)

tibble(
  Scenario               = names(scen),
  `Change (in)`          = b1 * scen,
  `As % of mean`         = 100 * b1 * scen / mean_precip,
  `As % of SD`           = 100 * b1 * scen / sd_precip
) |>
  knitr::kable(
    digits = 1,
    caption = "Effect sizes implied by Model 1. The last column is the one that matters for planning: it compares the ENSO signal against the year-to-year noise it must be detected within."
  )
```

**Practical significance.** The effect is real but small relative to the
noise a water manager actually faces. A one-standard-deviation swing in
winter ONI moves expected precipitation by
`r sprintf("%.1f", b1 * sd(d$oni_djf))` inches --
`r sprintf("%.0f", 100 * b1 * sd(d$oni_djf) / mean_precip)`% of the average
year, but only
`r sprintf("%.0f", 100 * b1 * sd(d$oni_djf) / sd_precip)`% of a standard
deviation of annual precipitation. Put the two extremes side by side: the
expected gap between a strong El Nino and a strong La Nina is about
`r sprintf("%.1f", b1 * 3)` inches, against a year-to-year standard
deviation of `r sprintf("%.1f", sd_precip)` inches. ENSO shifts the
distribution; it does not determine the outcome.

That asymmetry is the practical message. Table 2 makes it concrete: El Nino
years average `r sprintf("%.1f", mean(d$precip_in[d$enso_phase == "El Niño"]))`
inches against `r sprintf("%.1f", mean(d$precip_in[d$enso_phase == "La Niña"]))`
for La Nina years, yet the wettest La Nina year on record still exceeded
the average El Nino year. Planning that treats an El Nino forecast as a
guarantee of a wet year is not supported by 76 years of evidence.

## Conclusion

Across four specifications, winter ENSO intensity is positively associated
with California water-year precipitation. The central estimate is
`r sprintf("%.2f", slope_linear)` inches per degree of winter ONI
(equivalently about `r sprintf("%.0f", pct_per_oni)`% more precipitation
per degree), it is statistically significant at the 5% level under robust
standard errors, and it is unchanged by controls for the previous winter's
ENSO state and for trend.

The more useful finding, though, is how *little* of California's
hydrological variability this explains --
`r sprintf("%.1f", 100 * r2_linear)`% in Model 1. The historical record
contains very strong El Nino winters that produced ordinary years and
neutral winters that produced near-record ones. For the agencies that
motivated this question, the honest summary is that ENSO is a genuine but
weak tilt in the odds: useful as one input among many for water-supply
planning, and not sufficient on its own to anticipate a wet or dry year.

We make no causal claim. ENSO is not randomly assigned, our sample is a
single realization of one climate system observed over 76 years, and the
outcome is a statewide aggregate that averages over the strong regional
differences documented in prior work (Jong, Ting & Seager 2016). What this
report establishes is the shape and size of a historical association, which
is what a descriptive analysis can honestly deliver.


\newpage

## Appendix

### A.1 References

- Cayan, D. R., Redmond, K. T., and Dettinger, M. D. (1999). ENSO and hydrologic extremes in the western United States. *Journal of Climate*.
- Guirguis, K., et al. (2024). Reinterpreting ENSO's role in modulating impactful precipitation events in California. *Geophysical Research Letters*.
- Jong, B.-T., Ting, M., and Seager, R. (2016). El Niño's impact on California precipitation: seasonality, regionality, and El Niño intensity. *Environmental Research Letters*.
- NOAA National Centers for Environmental Information. *El Niño/Southern Oscillation (ENSO): Southern Oscillation Index (SOI) calculation.* <https://www.ncei.noaa.gov/access/monitoring/enso/soi> (accessed August 2026).
- Redmond, K. T., and Koch, R. W. (1991). Surface climate and streamflow variability in the western United States and their relationship to large-scale circulation indices. *Water Resources Research*.
- Schonher, T., and Nicholson, S. E. (1989). The relationship of California rainfall to ENSO events. *Journal of Climate*.

### A.2 Data Sources

The whole pipeline is reproducible from a single command, `Rscript src/data/make_analysis_data.R`, which rebuilds every input from source and writes the one file this report reads.

- **Precipitation.** The intended source was NOAA NCEI, *Climate at a Glance, Statewide Time Series*, as 12-month totals ending September (<https://www.ncei.noaa.gov/access/monitoring/climate-at-a-glance/statewide/time-series>). That service returned HTTP 503 throughout the analysis period, so `src/data/make_ca_wateryear_precip.R` falls back to a composite built from NOAA GHCN-Daily station records (`data/external/ca_monthly_rainfall.csv`), summing each station's twelve monthly totals per water year and averaging across stations. The script prefers the official series automatically whenever `data/external/data_rainfall.csv` is present, and the `precip_source` column in the analysis file records which was used.
- **Oceanic Niño Index (ONI).** Read directly from the NOAA Climate Prediction Center table at <https://www.cpc.ncep.noaa.gov/data/indices/oni.ascii.txt>, preserved raw at `data/external/oni_download.txt` and reduced to one row per water year by `src/data/make_oni_djf_wateryear.R` (`data/interim/oni_djf_wateryear.csv`). We read CPC directly rather than via the `rsoi` package because `rsoi` recomputes ONI as a rolling three-month mean and therefore cannot produce the DJF window for WY1950.
- **Southern Oscillation Index (SOI).** From the NOAA Climate Prediction Center, <https://www.cpc.ncep.noaa.gov/data/indices/soi>, preserved raw at `data/external/soi_download.txt` and reduced to the Nov–Mar reversed-SOI series by `src/data/make_soi_wateryear.R` (`data/interim/soi_ndjfm_wateryear.csv`).
- **GHCN-Daily station metadata**, the measurement network behind the precipitation composite, comes from NOAA NCEI, <https://www.ncei.noaa.gov/pub/data/ghcn/daily/>, with the California station list at `data/external/ca_station_list.csv`.
- **Merge and validation.** `src/data/make_analysis_data.R` joins the three interim files on water year and refuses to write unless the result is a complete consecutive run of 76 water years with no missing values in either analysis variable, mutually exclusive phase indicators, and exactly one missing lagged ONI (WY1950). It writes the canonical `data/processed/ca_enso_analysis.csv`.
- **Data dictionary.** `references/data_dictionary.csv` (rendered as `references/datadictionary.png`) documents every variable's type, units, range, source, and derivation, including all thirteen columns of the canonical analysis file.

```{=latex}
\newpage
\begin{landscape}
```

### A.5 Data Dictionary

```{r data-dictionary-table}
# The team's variable-level documentation (Data/data_dictionary.csv), rendered
# so the appendix is self-contained [assignment prompt - rubric: no output
# dumps]. format = "pipe" lets pandoc wrap the long description text in PDF.
read.csv(here("references", "data_dictionary.csv")) |>
  select(
    Variable = variable, Type = type, Units = units,
    `Range / levels` = range_or_levels, Description = description
  ) |>
  knitr::kable(
    format = "pipe",
    caption = "Data dictionary for the canonical analysis file (from Data/data_dictionary.csv)."
  )
```

```{=latex}
\end{landscape}
```

