---
title: "ENSO Intensity and California Statewide Precipitation: Water-Year Analysis"
subtitle: "DATASCI 203 — Group 2"
author: "Daniel Bowd, Gunhui Kim, Rishi Raj, Sasha Schaps"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: true
    fig_caption: true
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 3.4
)

library(here)
library(tidyverse)
library(sandwich)
library(lmtest)
library(stargazer)
library(car)
library(knitr)
library(kableExtra)
library(moments)

options(scipen = 999)

# PATHS (via here() so they resolve from the project root on any machine)
ONI_PATH      <- here("data", "interim", "oni_djf_wateryear.csv")
SOI_PATH      <- here("data", "interim", "soi_ndjfm_wateryear.csv")
# NOAA statewide water-year total. NOTE: the raw NOAA "Climate at a Glance"
# file this chunk was written for (skip=2, Date=YYYYMM) is not in the repo yet;
# data/external/california_annual_precipitation_1950_2025.csv has columns
# (year, annual_precip_inches) and needs a different reader. Repoint once the
# canonical statewide file lands in data/external/.
RAINFALL_PATH <- here("data", "external", "data_rainfall.csv")
REGIONAL_PATH <- here("data", "interim", "ca_rainfall_enso_model_data.csv")  # fallback proxy
```


# Data

## The ENSO series

```{r oni}
oni <- read_csv(ONI_PATH, col_types = cols(
  water_year   = col_integer(),
  oni_djf      = col_double(),
  enso_phase   = col_character(),
  enso_strength= col_character(),
  el_nino      = col_integer(),
  la_nina      = col_integer(),
  oni_djf_lag1 = col_double(),
  time_trend   = col_integer()
)) %>%
  mutate(
    enso_phase = factor(enso_phase,
                        levels = c("neutral", "el_nino", "la_nina"),
                        labels = c("Neutral", "El Nino", "La Nina"))
  )

soi <- read_csv(SOI_PATH, col_types = cols(
  water_year = col_integer(),
  rsoi       = col_double()
))
```

The ONI file is pre-built: the December-February mean, NOAA's phase and strength classifications, 0/1 phase indicators, a one-year lag, and a lineartime trend. No derivation is needed here, which removes a whole class of possible error - but it also means the water-year alignment is inherited rather than verified, so the checks below matter.

```{r oni-validate}
tibble(
  Check = c("Water years", "Range", "Mean ONI", "SD ONI",
            "Phase counts (La Nina / Neutral / El Nino)",
            "Lag available for"),
  Value = c(
    as.character(nrow(oni)),
    paste0("WY", min(oni$water_year), "-WY", max(oni$water_year)),
    sprintf("%.2f", mean(oni$oni_djf)),
    sprintf("%.2f", sd(oni$oni_djf)),
    paste(sum(oni$la_nina), sum(oni$enso_phase == "Neutral"),
          sum(oni$el_nino), sep = " / "),
    paste0(sum(!is.na(oni$oni_djf_lag1)), " years")
  )
) %>%
  kable(booktabs = TRUE, caption = "ENSO series validation.") %>%
  kable_styling(latex_options = "hold_position")

stopifnot(
  nrow(oni) == n_distinct(oni$water_year), #one row per water year
  !any(is.na(oni$oni_djf)),
  all(oni$el_nino + oni$la_nina <= 1) #phases mutually exclusive
)
```

## The precipitation series

The preferred outcome is NOAA's statewide water-year total. If that file is notpresent, the chunk below falls back to a station-weighted composite of the regional wet-season data, and flags the substitution loudly. **The fallback is
not the same construct** — it covers November through April, so it runs below the statewide annual figure (need to report).

```{r precip}
use_proxy <- !file.exists(RAINFALL_PATH)

if (!use_proxy) {

  # NOAA Climate at a Glance ships two comment lines before the header row:
  #   # California October-September Precipitation
  #   # Units: Inches
  #   Date,Value
  # Date is YYYYMM with MM = 09, the month the water year ends, so the first
  # four characters are the water year.
  precip <- read_csv(RAINFALL_PATH, skip = 2,
                     col_types = cols(Date = col_character(),
                                      Value = col_double())) %>%
    transmute(water_year = as.integer(substr(Date, 1, 4)),
              precip_in  = Value)

  stopifnot(all(substr(read_lines(RAINFALL_PATH, n_max = 2), 1, 1) == "#"))

} else {

  precip <- read_csv(REGIONAL_PATH, col_types = cols()) %>%
    mutate(cal_year = as.integer(format(ym, "%Y")),
           wy = if_else(month >= 10, cal_year + 1L, cal_year)) %>%
    group_by(region, wy) %>%
    summarise(mm = sum(prcp), stn = mean(n_stations), n_m = n(), .groups = "drop") %>%
    filter(n_m == 6) %>%
    group_by(water_year = wy) %>%
    summarise(precip_in = weighted.mean(mm / 25.4, stn),
              .groups = "drop")
}
```

```{r precip-flag, results='asis'}
if (use_proxy) {
  cat("\\begin{center}\\fbox{\\parbox{0.9\\textwidth}{\\textbf{Fallback outcome in use.} ",
      "The NOAA statewide series was not found, so results below use a ",
      "station-weighted regional wet-season composite. Correlation with the ONI ",
      "is 0.245 against 0.257 for the statewide series, so the pipeline is sound, ",
      "but levels and every intercept are not comparable to published statewide ",
      "figures.}}\\end{center}\n\n", sep = "")
}
```

## Merge

```{r merge}
df <- oni %>%
  inner_join(soi, by = "water_year") %>%
  inner_join(precip, by = "water_year") %>%
  arrange(water_year) %>%
  mutate(rsoi_lag1 = lag(rsoi)) %>%
  filter(!is.na(precip_in), !is.na(oni_djf_lag1), !is.na(rsoi_lag1)) %>%  # common sample, all models
  arrange(water_year)

stopifnot(nrow(df) == n_distinct(df$water_year), all(df$precip_in > 0))
```

All six models are estimated on these `r nrow(df)` water years. Restricting to
rows with a defined lag costs the first sample year but makes AIC and the nested
F-test legitimate comparisons rather than apples-to-oranges.


# Descriptive analysis

```{r descriptives}
tibble(
  Variable = c("Precipitation (in)", "ONI (Dec-Feb, C)",
               "Previous winter ONI (C)", "Reversed SOI (Nov-Mar)",
               "Previous winter reversed SOI"),
  N        = c(nrow(df), nrow(df), nrow(df), nrow(df), nrow(df)),
  Mean     = c(mean(df$precip_in), mean(df$oni_djf), mean(df$oni_djf_lag1), mean(df$rsoi), mean(df$rsoi_lag1)),
  SD       = c(sd(df$precip_in), sd(df$oni_djf), sd(df$oni_djf_lag1), sd(df$rsoi), sd(df$rsoi_lag1)),
  Min      = c(min(df$precip_in), min(df$oni_djf), min(df$oni_djf_lag1), min(df$rsoi), min(df$rsoi_lag1)),
  Median   = c(median(df$precip_in), median(df$oni_djf), median(df$oni_djf_lag1), median(df$rsoi), median(df$rsoi_lag1)),
  Max      = c(max(df$precip_in), max(df$oni_djf), max(df$oni_djf_lag1), max(df$rsoi), max(df$rsoi_lag1))
) %>%
  kable(booktabs = TRUE, digits = 2,
        caption = "Descriptive statistics for the analysis variables.") %>%
  kable_styling(latex_options = "hold_position")
```

```{r phase-summary}
df %>%
  group_by(`ENSO phase` = enso_phase) %>%
  summarise(`Water years` = n(),
            `Mean precip (in)` = mean(precip_in),
            `SD (in)` = sd(precip_in),
            `Driest` = min(precip_in),
            `Wettest` = max(precip_in),
            .groups = "drop") %>%
  kable(booktabs = TRUE, digits = 1,
        caption = "Precipitation by ENSO phase.") %>%
  kable_styling(latex_options = "hold_position")
```

```{r strength-summary}
df %>%
  mutate(enso_strength = factor(enso_strength,
           levels = c("neutral", "weak", "moderate", "strong", "very_strong"))) %>%
  group_by(Strength = enso_strength) %>%
  summarise(`Water years` = n(), `Mean precip (in)` = mean(precip_in),
            `SD (in)` = sd(precip_in), .groups = "drop") %>%
  kable(booktabs = TRUE, digits = 1,
        caption = "Precipitation by ENSO strength bin. The very strong bin holds
                   only a handful of years and cannot support inference on its own.") %>%
  kable_styling(latex_options = "hold_position")
```

```{r welch}
wt <- t.test(precip_in ~ el_nino, data = df)
cat(sprintf("Welch two-sample t-test, El Nino years vs all others: t = %.2f, df = %.1f, p = %.3f\n",
            wt$statistic, wt$parameter, wt$p.value))
```

## Figures

```{r fig-series, fig.cap="Statewide water-year precipitation, coloured by ENSO phase."}
ggplot(df, aes(x = water_year, y = precip_in, fill = enso_phase)) +
  geom_col() +
  scale_fill_manual(values = c("Neutral" = "#1baf7a",
                               "El Nino" = "#eb6834",
                               "La Nina" = "#2a78d6")) +
  labs(x = "Water year", y = "Precipitation (inches)", fill = "ENSO phase") +
  theme_minimal(base_size = 10) + theme(legend.position = "bottom")
```

```{r fig-scatter, fig.cap="Precipitation against winter ONI, with linear and quadratic fits."}
ggplot(df, aes(x = oni_djf, y = precip_in)) +
  geom_point(aes(colour = enso_phase, shape = enso_phase), size = 2, alpha = 0.85) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              colour = "#2a78d6", linewidth = 0.7) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE,
              colour = "#eb6834", linetype = "dashed", linewidth = 0.7) +
  scale_colour_manual(values = c("Neutral" = "#1baf7a",
                                 "El Nino" = "#eb6834",
                                 "La Nina" = "#2a78d6")) +
  labs(x = "Winter ONI (degrees C, Dec-Feb)", y = "Precipitation (inches)",
       colour = "ENSO phase", shape = "ENSO phase") +
  theme_minimal(base_size = 10) + theme(legend.position = "bottom")
```

```{r fig-hist, fig.cap="Precipitation is right-skewed; the log transform largely corrects it."}
bind_rows(
  df %>% transmute(value = precip_in,      scale = "Levels (inches)"),
  df %>% transmute(value = log(precip_in), scale = "Log inches")
) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 18, fill = "#2a78d6", alpha = 0.85) +
  facet_wrap(~ scale, scales = "free_x") +
  labs(x = NULL, y = "Water years") +
  theme_minimal(base_size = 10)
```


# Models

```{r models}
m1 <- lm(precip_in      ~ oni_djf,                                data = df)
m2 <- lm(precip_in      ~ el_nino + la_nina,                      data = df)
m3 <- lm(precip_in      ~ oni_djf + oni_djf_lag1 + time_trend,    data = df)
m4 <- lm(log(precip_in) ~ oni_djf,                                data = df)
m5 <- lm(precip_in      ~ oni_djf + I(oni_djf^2),                 data = df)
m6 <- lm(precip_in      ~ oni_djf + rsoi + el_nino,               data = df)

models <- list(m1, m2, m3, m4, m5, m6)
rob_se <- lapply(models, function(m) sqrt(diag(vcovHC(m, type = "HC1"))))
```

| Model | Question |
|-------|----------|
| M1 | Is there a linear ENSO-precipitation association? |
| M2 | What does collapsing intensity into phase categories cost? |
| M3 | Is the association confounded by last winter's ENSO or by trend? |
| M4 | Is the relationship proportional rather than additive? |
| M5 | Does the relationship steepen at extreme ONI values? |
| M6 | Does nesting the continuous rSOI add anything beyond the baseline? |

```{r model-table, results='asis'}
stargazer(
  m1, m2, m3, m4, m5, m6,
  se               = rob_se,
  type             = "latex",
  header           = FALSE,
  title            = "ENSO intensity and California water-year precipitation.",
  column.labels    = c("M1", "M2", "M3", "M4", "M5", "M6"),
  model.numbers    = FALSE,
  covariate.labels = c("Winter ONI", "El Nino", "La Nina",
                       "Winter ONI squared", "Previous winter ONI", "Time trend",
                       "Reversed SOI"),
  omit.stat        = c("f", "ser"),
  notes            = "HC1 heteroskedasticity-robust standard errors.",
  notes.align      = "l",
  font.size        = "small"
)
```

## Practical significance

```{r practical}
b1 <- coef(m1)["oni_djf"]
tibble(
  Scenario = c("One SD increase in winter ONI",
               "Strong El Nino (ONI = +1.5)",
               "Very strong El Nino (ONI = +2.3)",
               "Strong La Nina (ONI = -1.5)"),
  `Change vs mean (in)` = c(b1 * sd(df$oni_djf), b1 * 1.5, b1 * 2.3, b1 * -1.5),
  `As pct of mean` = 100 * c(b1 * sd(df$oni_djf), b1 * 1.5,
                             b1 * 2.3, b1 * -1.5) / mean(df$precip_in),
  `As pct of SD`   = 100 * c(b1 * sd(df$oni_djf), b1 * 1.5,
                             b1 * 2.3, b1 * -1.5) / sd(df$precip_in)
) %>%
  kable(booktabs = TRUE, digits = 2,
        caption = "Effect sizes implied by Model 1. The final column is the one
                   that matters for planning: it compares the ENSO signal to the
                   year-to-year noise it must be detected against.") %>%
  kable_styling(latex_options = "hold_position")
```

## Model comparison

```{r comparison}
tibble(
  Model = c("M1 linear", "M2 phase dummies", "M3 lag + trend",
            "M4 log outcome", "M5 quadratic", "M6 ONI + SOI + El Nino"),
  `R-squared`   = map_dbl(models, ~ summary(.x)$r.squared),
  `Adj R-squared` = map_dbl(models, ~ summary(.x)$adj.r.squared),
  AIC           = map_dbl(models, AIC)
) %>%
  kable(booktabs = TRUE, digits = 4,
        caption = "M4's outcome is on a different scale, so its R-squared and AIC
                   are NOT comparable to the others. Compare it only to itself.") %>%
  kable_styling(latex_options = "hold_position")

anova(m1, m3)   # do the lag and trend jointly add anything?
anova(m1, m5)   # is the quadratic term warranted?
```


# Regression assumptions

```{r assumptions}
bg <- bgtest(m1, order = 2)
bp_lvl <- bptest(m1); bp_log <- bptest(m4)
e <- residuals(m1)

tibble(
  Assumption = c("IID / no serial correlation", "IID / no serial correlation",
                 "Homoskedasticity (levels)", "Homoskedasticity (log)",
                 "Normal errors (levels)", "Normal errors (log)",
                 "Skewness (levels)", "Skewness (log)"),
  Test  = c("Breusch-Godfrey(2)", "Lag-1 residual autocorrelation",
            "Breusch-Pagan", "Breusch-Pagan",
            "Shapiro-Wilk", "Shapiro-Wilk", "-", "-"),
  Value = c(sprintf("LM = %.2f, p = %.3f", bg$statistic, bg$p.value),
            sprintf("%.3f", cor(head(e, -1), tail(e, -1))),
            sprintf("p = %.3f", bp_lvl$p.value),
            sprintf("p = %.3f", bp_log$p.value),
            sprintf("p = %.3f", shapiro.test(residuals(m1))$p.value),
            sprintf("p = %.3f", shapiro.test(residuals(m4))$p.value),
            sprintf("%.2f", skewness(df$precip_in)),
            sprintf("%.2f", skewness(log(df$precip_in))))
) %>%
  kable(booktabs = TRUE, caption = "Classical linear model assumption checks.") %>%
  kable_styling(latex_options = "hold_position")
```

```{r vif-table}
vif(m3) %>% enframe(name = "Term", value = "VIF") %>%
  kable(booktabs = TRUE, digits = 2,
        caption = "Variance inflation factors, M3.") %>%
  kable_styling(latex_options = "hold_position")
```

```{r diag-plots, fig.height=5, fig.cap="Diagnostic plots for Models 1 and 4."}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(fitted(m1), residuals(m1), pch = 20, xlab = "Fitted", ylab = "Residuals",
     main = "M1 residuals"); abline(h = 0, lty = 2)
qqnorm(residuals(m1), pch = 20, main = "M1 Q-Q"); qqline(residuals(m1))
plot(fitted(m4), residuals(m4), pch = 20, xlab = "Fitted", ylab = "Residuals",
     main = "M4 residuals"); abline(h = 0, lty = 2)
qqnorm(residuals(m4), pch = 20, main = "M4 Q-Q"); qqline(residuals(m4))
par(mfrow = c(1, 1))
```

```{r acf, fig.height=2.8, fig.cap="Residual autocorrelation, Model 1."}
acf(residuals(m1), main = "")
```


# Reproducibility check

```{r repro}
tribble(
  ~Quantity,                       ~Expected,
  "Descriptives n = 76 (WY1950-2025)", "mean 22.18, SD 6.75, min 10.86, median 20.80, max 40.41",
  "Correlation with ONI",              "0.257",
  "Skewness levels -> log",            "0.65 -> 0.11",
  "Phase groups (La Nina/Neutral/El Nino)", "29 / 20 / 27",
  "Welch t-test, El Nino vs rest",     "t = 1.77, p = 0.084",
  "Model sample",                      "n = 75 (WY1951-2025)",
  "M1 slope",                          "1.645 (HC1 SE 0.807, p = 0.042)",
  "M1 intercept / R-squared",          "22.177 / 0.0620",
  "M2 El Nino / La Nina",              "3.251 (p = 0.120) / 0.348 (p = 0.840)",
  "M2 R-squared",                      "0.0477 (below M1)",
  "M3 ONI / lag / trend",              "1.765 (p = 0.028) / 1.379 (p = 0.047) / -0.027 (p = 0.425)",
  "M4 semi-elasticity",                "0.062 (p = 0.061)",
  "M5 quadratic term",                 "1.011 (p = 0.122)",
  "Nested F, M1 vs M3",                "2.03, p = 0.139",
  "Nested F, M1 vs M5",                "2.81, p = 0.098",
  "Breusch-Godfrey(2)",                "LM = 1.50, p = 0.471",
  "Lag-1 residual autocorrelation",    "-0.098",
  "Breusch-Pagan (levels / log)",      "p = 0.010 / p = 0.170",
  "Shapiro-Wilk (levels / log)",       "p = 0.009 / p = 0.043"
) %>%
  kable(booktabs = TRUE,
        caption = "Values computed independently in Python from the same two
                   source files. The descriptive rows also match the published
                   report exactly, which confirms the pipeline reproduces the
                   original analysis. If the knit disagrees with any row here,
                   something drifted.") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

```{r session}
sessionInfo()
```
